Disease Knowledge Transfer across Neurodegenerative Diseases

### 2019-10-08 paper/pres/poster-dkt.tex
2nd rehearsal DKT

\author{\LARGE{R\u{a}zvan Marinescu\inst{1,2}, Marco Lorenzi\inst{3}, Stefano Blumberg\inst{1}, Alexandra Young\inst{1}, Pere Morell\inst{1}},\\ \LARGE{Neil Oxtoby\inst{1}, Arman Eshaghi\inst{1,4}, Keir Yong\inst{5}, Sebastian Crutch\inst{5}, Polina Golland\inst{2}, Daniel Alexander\inst{1}}\\
\begin{tabular}{l p{1cm} l p{1cm} l}
\large{$^1$Centre for Medical Image Computing, UCL, UK}  & & \large{$^3$University of C\^{o}te d'Azur, Inria Sophia Antipolis, France} & & \large{$^5$Dementia Research Centre, UCL, UK}\\
\large{$^2$Computer Science and Artificial Intelligence Laboratory, MIT, USA} & & \large{$^4$Queen Square MS Centre, UCL Institute of Neurology, UK} & & 
 \\
\end{tabular}
}

@@

\begin{document}
							%Align with edge of page, not margin
% \vspace{2cm}
% \includegraphics[scale=1,trim=0 0 0 200, clip]{Black_Landscape.pdf}
\hspace{-1cm}	
\begin{minipage}{50cm}					%Minipage for title contents
% \vspace{-18cm}
\maketitle
\end{minipage}
\begin{textblock}{0}(12.2,0.2)
\begin{tabular}{l}
\includegraphics[height=5.5cm]{mit_logo}\includegraphics[height=6cm]{qr_dkt.png}\\
\includegraphics[height=5cm]{ucl_logo}\\
\end{tabular}

@@

\newcommand{\lp}{\lambda_{d_i}^{\psi(k)}}
\newcommand{\lpuu}{\lambda_{d_i}^{\psi(k),(u)}}
\newcommand{\lpum}{\lambda_{d_i}^{\psi(k),(u-1)}}

@@

\hrule

@@

\heading{Aim} Infer progression of non-MRI biomarkers in rare neurodegenerative diseases (NDs) by leveraging larger datasets of common NDs.

\heading{Why} Posterior Cortical Atrophy (PCA): progression of non-MRI biomarkers not known $\rightarrow$ Identify outcome measures and suitable subjects for PCA clinical trials

@@

\heading{Challenges}

% \begin{itemize}
%  \item Datasets on rare neurodegenerative diseases (e.g., Posterior Cortical Atrophy) are unimodal (MRI only), cross-sectional and small.
%  \item The continuous progression of non-MRI markers in rare neurodegenerative diseases is not well understood
% \end{itemize}

\begin{figure}[H]
\fnt{
\begin{subfigure}{0.48\columnwidth}
  Typical Neurodeg. Diseases

  \begin{itemize}
   \item Large datasets \yes
   \item Multimodal imaging \yes
   \item Longitudinal \yes
%    \item Well-understood \yes

  \end{itemize}
\end{subfigure}
\begin{subfigure}{0.48\columnwidth}

   Rare Neurodeg. Diseases

  \begin{itemize}
   \item Small datasets \no
   \item MRI only \no
   \item Cross-sectional \no
%    \item Poorly understood \no
  \end{itemize}

\end{subfigure}
 }
\end{figure}

@@

\heading{Method}

@@

  \end{subfigure}
  \begin{subfigure}{0.49\columnwidth}
   \centering
%    \vspace{-1.2em}
   \fnt{2. Dysfunction trajectory modelled using region-specific biomarkers}\\
   $ y_{ijk} = g( \gamma_{ij}^{\psi(k)} ; \theta_k) + N(0,\epsilon_k) $
%    \vspace{0.6em}
%    \includegraphics[width=\textwidth]{dysfuncEvol}
  \end{subfigure}

@@

\heading{Inference:} Loopy belief propagation

@@

\begin{figure}[H]
\begin{algorithm}[H]
\algoFnt
 Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
  \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
   \tcp*[l]{\algoFnt{Estimate biomarker trajectories (disease agnostic)}}
%     \For{$k=1$ to $K$}{
      ${\theta_k\uu = \argmin_{\theta_k} \textcolor{red}{\sum_{(i,j) \in \Omega_k}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2 }$\\
%       ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
%     }
     \tcp*[l]{\algoFnt{Estimate dysfunction trajectories (disease specific)}} 
%     \For{$d=1 \in \mathbb{D}$}{
%       \For{$l=1 \in \Lambda$}{
        ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \textcolor{red}{\sum_{(i,j,k) \in \Omega_{d,l}}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
        \right) \right]^2 }$\\
%       }
%     }
    \tcp*[l]{\algoFnt{Estimate subject-specific time shifts}} 
%     \For{$i=1 \in [1, \dots, S]$}{
      ${\beta_i\uu = \argmin_{\beta_i} \textcolor{red}{\sum_{(j,k) \in \Omega_i}} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
      \right) \right]^2 }$\\
%     }
}
% \normalfont
\end{algorithm}
% \caption[The algorithm for estimating the DKT parameters]{The algorithm used to estimate the DKT parameters, based on loopy belief-propagation.}
\label{fig:dktAlgo}
\end{figure}

\heading{Datasets}

\begin{itemize}
 \item Dementia Research Center cohort: MRI scans from 76 PCA, 67 tAD, 87 controls for training, 10 PCA with DTI for validation.
 \item TADPOLE dataset (ADNI) split into three subgroups with different progressions: 21 hippocampal, 35 cortical, 27 subcortical.
 \item Synthetic dataset mimicking the cohorts above: 50 subjects with "synthetic PCA", 100 subjects with "synthetic AD".
\end{itemize}

% The cohort from the Dementia Research Centre UK used for validation consisted of 10 subjects diagnosed with Posterior Cortical Atrophy, with a mean age of 59.4, 40\% females, as well as 10 age-matched controls with a mean age of 59.3, 50\% females.
% 
% 
% For the validation on TADPOLE subgroups, we used applied the SuStaIn model on TADPOLE to split the population into three subgroups with different progression: hippocampal, cortical and subcortical subypes with prominent atrophy in the hippocampus, cortical and subcortical areas respectively.
% The resulting subgroups had the following demographics:
% 
% \begin{table}
%  \begin{tabular}{c | c c c c}
% \textbf{Cohort} & \textbf{Nr. subjects} & \textbf{Nr. visits} & \textbf{Age (baseline)} & \textbf{Gender (\%F)}\\
% \hline
% Controls (Hippocampal) & 31 & 2.3 $\pm$ 1.8 & 74.4 $\pm$ 6.9 & 38\%\\
% AD (Hippocampal) & 21 & 1.5 $\pm$ 0.8 & 74.5 $\pm$ 5.5 & 42\%\\
% \hline
% Controls (cortical) & 21 & 2.3 $\pm$ 1.3 & 70.9 $\pm$ 5.4 & 42\%\\
% AD (cortical) & 35 & 1.7 $\pm$ 0.9 & 72.8 $\pm$ 7.4 & 28\%\\
% \hline
% Controls (subcortical) & 28 & 3.0 $\pm$ 1.5 & 73.7 $\pm$ 6.5 & 42\%\\
% AD (subcortical) & 27 & 1.6 $\pm$ 0.9 & 73.7 $\pm$ 7.5 & 33\%\\
%  \end{tabular}
%  \caption{Demographics of the subjects in the three TADPOLE subgroups.}
%  \label{tab:demogTadSubtypes}
% \end{table}

\heading{Results}

\vspace{0.6em}
\begin{itemize}
 \item \fnt{In synthetic experiment, the estimated parameters are close to the true parameters.}
\begin{figure}[H]
\includegraphics[width=\columnwidth]{../figures/compTrueParams105_synth1_JMD.pdf}
  \label{fig:dktSynthTrajCompTrue}
\end{figure}

@@

\item \fnt{Our model has favourable performance compared to other models, on two different datasets.}
\vspace{0.3em}
\begin{figure}[H]
\fontsize{27}{30}\selectfont
\begin{tabular}{c | c c c c c c}
\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippocam.} & \textbf{Occipital} & \textbf{Parietal} & \textbf{Temporal}\\
& \multicolumn{6}{c}{\textbf{TADPOLE: Hippocampal subgroup to Cortical subgroup}}\\
DKT (ours) &      0.56 $\pm$ 0.23 &    \textbf{0.35 $\pm$ 0.17} &        \textbf{0.58 $\pm$ 0.14} &     -0.10 $\pm$ 0.29 &     \textbf{0.71 $\pm$ 0.11} &     \textbf{0.34 $\pm$ 0.26} \\
Latent stage &      0.44 $\pm$ 0.25 &    0.34 $\pm$ 0.21 &       0.34 $\pm$ 0.24* &     \textbf{-0.07 $\pm$ 0.22} &     0.64 $\pm$ 0.16 &    0.08 $\pm$ 0.24* \\
Multivariate &      \textbf{0.60 $\pm$ 0.18} &   0.11 $\pm$ 0.22* &       0.12 $\pm$ 0.29* &     -0.22 $\pm$ 0.22 &   -0.44 $\pm$ 0.14* &   -0.32 $\pm$ 0.29* \\
Spline &    -0.24 $\pm$ 0.25* &  -0.06 $\pm$ 0.27* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.10 $\pm$ 0.25* \\
Linear &    -0.24 $\pm$ 0.25* &   0.20 $\pm$ 0.25* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.13 $\pm$ 0.23* \\
& \multicolumn{6}{c}{\textbf{typical Alzheimer's to Posterior Cortical Atrophy}}\\
DKT (ours) &    0.77 $\pm$ 0.11 &    0.39 $\pm$ 0.26 &      0.75 $\pm$ 0.09 &    0.60 $\pm$ 0.14 &    \textbf{0.55 $\pm$ 0.24} &    \textbf{0.35 $\pm$ 0.22} \\
Latent stage &    \textbf{0.80 $\pm$ 0.09} &    \textbf{0.53 $\pm$ 0.17} &      \textbf{0.80 $\pm$ 0.12} &    0.56 $\pm$ 0.18 &    0.50 $\pm$ 0.21 &    0.32 $\pm$ 0.24 \\
Multivariate &   0.73 $\pm$ 0.09 &   0.45 $\pm$ 0.22  &    0.71 $\pm$ 0.08 & -0.28 $\pm$ 0.21* &  0.53 $\pm$ 0.22  &  0.25 $\pm$ 0.23* \\
Spline &   0.52 $\pm$ 0.20* &  -0.03 $\pm$ 0.35* &     0.66 $\pm$ 0.11* &   0.09 $\pm$ 0.25* &    0.53 $\pm$ 0.20 &   0.30 $\pm$ 0.21* \\
Linear &   0.52 $\pm$ 0.20* &    0.34 $\pm$ 0.27 &     0.66 $\pm$ 0.11* &    \textbf{0.64 $\pm$ 0.17} &    0.54 $\pm$ 0.22 &   0.30 $\pm$ 0.21* \\
\end{tabular}
\vspace{0.5em}
% \caption[Performance evaluation of DKT and other models]{Performance evaluation of DKT and four other statistical models of decreasing complexity. }
% \end{footnotesize}
\label{sec:dktPerfMetrics}
\end{figure}

@@

\heading{Conclusion}

>>>>

% \author{\LARGE{R\u{a}zvan Marinescu\inst{1,2}, Marco Lorenzi\inst{3}, Stefano Blumberg\inst{1}, Alexandra Young\inst{1}, Pere Morell\inst{1}},\\ \LARGE{Neil Oxtoby\inst{1}, Arman Eshaghi\inst{1,4}, Keir Yong\inst{5}, Sebastian Crutch\inst{5}, Polina Golland\inst{2}, Daniel Alexander\inst{1}}\\}

\author{\LARGE{R\u{a}zvan V. Marinescu, Marco Lorenzi, Stefano B. Blumberg, Alexandra L. Young, Pere Morell},\\ \LARGE{Neil Oxtoby, Arman Eshaghi, Keir Yong, Sebastian Crutch, Polina Golland, Daniel C. Alexander}\\}

% \begin{tabular}{l p{1cm} l p{1cm} l}
% \large{$^1$Centre for Medical Image Computing, UCL, UK}  & & \large{$^3$University of C\^{o}te d'Azur, Inria Sophia Antipolis, France} & & \large{$^5$Dementia Research Centre, UCL, UK}\\
% \large{$^2$Computer Science and Artificial Intelligence Laboratory, MIT, USA} & & \large{$^4$Queen Square MS Centre, UCL Institute of Neurology, UK} & & 
%  \\
% \end{tabular}

@@

\begin{document}
							%Align with edge of page, not margin
% \vspace{2cm}
% \includegraphics[scale=1,trim=0 0 0 200, clip]{Black_Landscape.pdf}
\hspace{-1cm}	
\begin{minipage}{50cm}					%Minipage for title contents
% \vspace{-18cm}
\maketitle
\end{minipage}
\begin{textblock}{0}(9.95,0.2)
\begin{tabular}{l}
\includegraphics[height=5.5cm]{mit_logo}\includegraphics[height=5.5cm]{inria_logo}\includegraphics[height=6cm]{qr_dkt.png}\\
\hspace{11.45cm}\includegraphics[height=5cm]{ucl_logo}\\
\end{tabular}

@@

\newcommand{\lp}{\lambda_{d_i}^{k}}
\newcommand{\lpuu}{\lambda_{d_i}^{k,(u)}}
\newcommand{\lpum}{\lambda_{d_i}^{k,(u-1)}}

@@

\vspace{0.2em}
\hrule
\vspace{0.2em}

@@

\heading{Aim}\;\;\;\;Infer progression of multimodal biomarkers in rare neurodegenerative diseases (NDs) by leveraging larger datasets of common NDs.

\heading{Why}\;\;\;\;Posterior Cortical Atrophy (PCA): progression of \underline{multimodal} biomarkers not known $\rightarrow$ Identify outcome measures and suitable subjects for PCA clinical trials

@@

% \heading{Challenges}
% 
% \begin{figure}[H]
% \fnt{
% 
% \begin{subfigure}{0.48\columnwidth}
%  
%    Rare Neurodeg. Diseases
%   
%   \begin{itemize}
%    \item Small datasets \no
%    \item MRI only \no
%    \item Cross-sectional \no
% %    \item Poorly understood \no
%   \end{itemize}
%  
% \end{subfigure}
%  }
% \end{figure}

\begin{figure}[H]
 \centering
 \includegraphics[width=\columnwidth]{DKT_data_availability}
\end{figure}

\end{multicols}
\hrule
\vspace{0.2em}

@@

\heading{1. Intuition}

\begin{itemize}
 \item Diseases affect different brain regions un-equally, but underlyining mechanisms are the same (amyloid cascade)
 \item \textbf{Idea:} each brain region follows ``its own disease course'', common across diseases $\rightarrow$ dysfunction score
\end{itemize}

\begin{figure}[H]

\includegraphics[width=\columnwidth]{DKT_intuition}

\end{figure}

\heading{2. Method}

@@

  \end{subfigure}
  \begin{subfigure}{0.49\columnwidth}
   \centering
%    \vspace{-1.2em}
   \fnt{2. Dysfunction trajectory modelled using region-specific biomarkers}\\
   $ y_{ijk} = g( \gamma_{ij}^{k} ; \theta_k) + N(0,\epsilon_k) $
%    \vspace{0.6em}
%    \includegraphics[width=\textwidth]{dysfuncEvol}
  \end{subfigure}

@@

\heading{3. Inference with belief propagation}

@@

% \begin{itemize}
%  \item Uses loopy belief propagation
% \end{itemize}

\begin{figure}[H]
\begin{algorithm}[H]
\algoFnt{
%  Initialise $\boldsymbol{\theta}^{(0)}$, $\boldsymbol{\lambda}^{(0)}$, $\boldsymbol{\beta}^{(0)}$\\
  \While{$\boldsymbol{\theta}$, $\boldsymbol{\lambda}$, $\boldsymbol{\beta}$ not converged}{
   \tcp*[l]{\algoFnt{Estimate biomarker trajectories (disease agnostic)}}
%     \For{$k=1$ to $K$}{
      ${\theta_k\uu = \argmin_{\theta_k} \textcolor{red}{\sum_{(i,j) \in \Omega_k}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\right) \right]^2 }$\\
%       ${\epsilon_k\uu = \frac{1}{|\Omega_k|} \sum_{(i,j) \in \Omega_k}    \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lpum) ; \theta_k\uu \right) \right]^2 }$\\
%     }
     \tcp*[l]{\algoFnt{Estimate dysfunction trajectories (disease specific)}} 
%     \For{$d=1 \in \mathbb{D}$}{
%       \For{$l=1 \in \Lambda$}{
        ${\lambda_{d}^{l, (u)} = \argmin_{\lambda_{d}^{l}} \textcolor{red}{\sum_{(i,j,k) \in \Omega_{d,l}}} \left[y_{ijk} - g\left(f(\beta_i\um + m_{ij}; \lambda_{d}^{l}) ; \theta_k\uu 
        \right) \right]^2 }$\\
%       }
%     }
    \tcp*[l]{\algoFnt{Estimate subject-specific time shifts}} 
%     \For{$i=1 \in [1, \dots, S]$}{
      ${\beta_i\uu = \argmin_{\beta_i} \textcolor{red}{\sum_{(j,k) \in \Omega_i}} \left[y_{ijk} - g\left(f(\beta_i + m_{ij}; \lpuu) ; \theta_k\uu
      \right) \right]^2 }$\\
%     }
}
}
% \normalfont
\end{algorithm}
% \caption[The algorithm for estimating the DKT parameters]{The algorithm used to estimate the DKT parameters, based on loopy belief-propagation.}
\label{fig:dktAlgo}
\end{figure}

\heading{4. Datasets}

\begin{itemize}
 \item Dementia Research Center cohort: MRI scans from 76 PCA, 67 tAD, 87 controls for training, 10 PCA with DTI for validation.
 \item TADPOLE dataset (ADNI) split into three subgroups with different progressions: 21 hippocampal, 35 cortical, 27 subcortical.
%  \item Synthetic dataset mimicking the cohorts above: 50 subjects with "synthetic PCA", 100 subjects with "synthetic AD".
\end{itemize}

\vspace{0.6em}
\heading{5. Results}

% \vspace{0.6em}
\begin{itemize}
%  \item \fnt{In synthetic experiment, the estimated parameters are close to the true parameters.}
% \begin{figure}[H]
% \includegraphics[width=\columnwidth]{../figures/compTrueParams105_synth1_JMD.pdf}
%   \label{fig:dktSynthTrajCompTrue}
% \end{figure}

@@

\item \fnt{Our model has favourable performance compared to other models, on two different datasets.}
\vspace{0.3em}
\begin{figure}[H]
\fontsize{27}{30}\selectfont
\begin{tabular}{c | c c c c c c}
\textbf{Model} & \textbf{Cingulate} & \textbf{Frontal} & \textbf{Hippocam.} & \textbf{Occipital} & \textbf{Parietal} & \textbf{Temporal}\\
& \multicolumn{6}{c}{\textbf{TADPOLE: Hippocampal subgroup to Cortical subgroup}}\\
DKT (ours) &      0.56 $\pm$ 0.23 &    \textbf{0.35 $\pm$ 0.17} &        \textbf{0.58 $\pm$ 0.14} &     -0.10 $\pm$ 0.29 &     \textbf{0.71 $\pm$ 0.11} &     \textbf{0.34 $\pm$ 0.26} \\
AD model &      0.44 $\pm$ 0.25 &    0.34 $\pm$ 0.21 &       0.34 $\pm$ 0.24* &     \textbf{-0.07 $\pm$ 0.22} &     0.64 $\pm$ 0.16 &    0.08 $\pm$ 0.24* \\
Multivariate &      \textbf{0.60 $\pm$ 0.18} &   0.11 $\pm$ 0.22* &       0.12 $\pm$ 0.29* &     -0.22 $\pm$ 0.22 &   -0.44 $\pm$ 0.14* &   -0.32 $\pm$ 0.29* \\
Spline &    -0.24 $\pm$ 0.25* &  -0.06 $\pm$ 0.27* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.10 $\pm$ 0.25* \\
Linear &    -0.24 $\pm$ 0.25* &   0.20 $\pm$ 0.25* &        0.58 $\pm$ 0.17 &     -0.16 $\pm$ 0.27 &    0.23 $\pm$ 0.25* &    0.13 $\pm$ 0.23* \\
& \multicolumn{6}{c}{\textbf{typical Alzheimer's to Posterior Cortical Atrophy}}\\
DKT (ours) &    0.77 $\pm$ 0.11 &    0.39 $\pm$ 0.26 &      0.75 $\pm$ 0.09 &    0.60 $\pm$ 0.14 &    \textbf{0.55 $\pm$ 0.24} &    \textbf{0.35 $\pm$ 0.22} \\
AD model &    \textbf{0.80 $\pm$ 0.09} &    \textbf{0.53 $\pm$ 0.17} &      \textbf{0.80 $\pm$ 0.12} &    0.56 $\pm$ 0.18 &    0.50 $\pm$ 0.21 &    0.32 $\pm$ 0.24 \\
Multivariate &   0.73 $\pm$ 0.09 &   0.45 $\pm$ 0.22  &    0.71 $\pm$ 0.08 & -0.28 $\pm$ 0.21* &  0.53 $\pm$ 0.22  &  0.25 $\pm$ 0.23* \\
Spline &   0.52 $\pm$ 0.20* &  -0.03 $\pm$ 0.35* &     0.66 $\pm$ 0.11* &   0.09 $\pm$ 0.25* &    0.53 $\pm$ 0.20 &   0.30 $\pm$ 0.21* \\
Linear &   0.52 $\pm$ 0.20* &    0.34 $\pm$ 0.27 &     0.66 $\pm$ 0.11* &    \textbf{0.64 $\pm$ 0.17} &    0.54 $\pm$ 0.22 &   0.30 $\pm$ 0.21* \\
\end{tabular}
\vspace{0.5em}
% \caption[Performance evaluation of DKT and other models]{Performance evaluation of DKT and four other statistical models of decreasing complexity. }
% \end{footnotesize}
\label{sec:dktPerfMetrics}
\end{figure}

@@

\heading{6. Conclusion}
